Synthesizing Quotient Lenses

### 2018-03-09 papers/QuotientLenses/main.tex
Checkpoint polishing and comments

In our previous work in~\cite{optician}, we presented a solution to these
shortcomings of bijective lenses by developing a tool called {\em Optician}
which enables a programmer to synthesize a bijective lens from a pair of regular
expressions $(S, T)$ and a set of input-output example \cd{exs} pairs that
determines how the synthesized lens should behave on those examples. With {\em
Optician}, the programmer need only spend the effort of specifying the data
formats and the behaviour of the lens on a few inputs in order to define a
bijective lens. As a result defining bijective lenses becomes faster and less
error prone, since the programmer need only invoke the synthesis procedure
\lstinline{synth S <=> T using exs} in order to define a bijective lens rather
than manually derive the lens herself.

@@

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $c$ enables us to express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(c)$ and $K(c)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonizer(R) &=& id_{\mathcal{L}(R)} \\
\canonizer(\project{R}{s})(w) &=& s\\
\canonizer(\squash{c_1}{c_2}{f})(w) &=&
\begin{cases}
\canonizer(c_1)(f(w)) & \text{if } w \in \mathcal{L}(W(c_1))\\
\canonizer(c_2)(w) & \text{otherwise}
\end{cases}\\
\canonizer(\normalize{R}{R'}{f}) &=& f\\
\canonizer(c_1 \; ; \; c_2) &=& \canonizer(c_2) \circ \canonizer(c_1)\\
\canonizer(c_1 \cdot c_2) &=& \canonizer(c_1) \cdot \canonizer(c_2)\\
\canonizer(c_1 \sep c_2)(w) &=&
\begin{cases}
\canonizer(c_1)(w) & \text{if } w \in \mathcal{L}(W(c_1))\\
\canonizer(c_2)(w) & \text{if } w \in \mathcal{L}(W(c_2))\\
\end{cases}\\
\canonizer(c^*) &=& \canonizer(c)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \textit{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\noindent which as we have explained is defined so that the strings that match
$c_i, c$ are placed in the canonical permutation which is the identity
permutation before $\canonizer(c_i)$ and $\canonizer(c)$ are applied
respectively.

Finally, Figure ~\ref{fig:relations} gives the induction defintion of the
equivalence relation $\eqrel{c}$. These are the set-theoretic semantics of QREs
as an equivalence relation on a regular language.

\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}
w \; \equiv_R \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' \text{ for all }w, w'\\
w \; \equiv_{\squash{c_1}{c_2}{f}} \; w' &\iff& f(w) \equiv_{c_2} w'
\text{ or } w \equiv_{c_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{c_1 \; ; \; c_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(c_2)) \text{ such that } w \; \equiv_{c_1} \; k, \; w' \;
\equiv_{c_1} \; k', \text{ and } k \; \equiv_{c_2} \; k'\\
w \; \equiv_{c_1 \cdot c_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{c_1}
\; {r'}_1, \; r_2 \; \equiv{c_n} \; {r'}_2\\
w \; \equiv_{c_1 \sep c_2} \; w' &\iff& w \; \equiv_{c_1} \; w'
\text{ or } \; w \; \equiv_{c_2} \; w'\\
w \; \equiv_{c^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{c} \; {r'}_i
\\
w \; \equiv_{\perm{c_1, \ldots, c_n}{c}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{ for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{c} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Well Formed QREs}
\subsubsection{Completeness of {\em normalize}}
Before we present the full inference rules, for QREs, we discuss the {\em
normalize} QRE which is special in two main ways, the first of which has to do
with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

@@

More formally, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$. We say
that a regular expression $R$ is \textit{strongly unambiguous}\cite{Sippu1988}
if and only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1,
S_2$ strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \sep S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

@@

Figure~\ref{fig:qrerules} gives the inferrence rules for deriving well-formed
QREs.
\begin{figure}[tp!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\RightLabel{(Id)}
\UnaryInfC{$\wf{\mathit{id}(R)}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\RightLabel{(Project)}
\UnaryInfC{$\wf{\project{R}{s}}$}
\DisplayProof
\hskip 1.5em
\end{center}

@@

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$W(c_1) \cap W(c_2) = \varnothing$}
\RightLabel{(Union)}
\BinaryInfC{$\wf{c_1 \sep c_2}$}
\DisplayProof
\
\end{center}
\end{center}
\caption{Well-formed QREs}
\label{fig:qrerules}
\end{figure}
As we already mentioned, the unambiguity conditions come into play when defining
QREs using the regular combinators. For example, the (Concat) inferrence rule
for concatenation says that the concatenation $c_1 \cdot c_2$ of QREs $c_1$ and
$c_2$ is well formed only if the concatentions of $W(c_1)$ and $W(c_2)$, and
$K(c_1)$ and $K(c_2)$ are unambiguous.

The most complicated inference rule is the (Perm) rule for the
\textit{permutation} combinator. The second hypothesis for the
\textit{permutation} rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c_{\sigma(n)})$ and $W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot
W(c_{\theta(n)})$ must be disjoint. This is because any string that matches the
\textit{permutation} combinator matches the language $W(c_{\theta(1)}) \cdot
W(c) \cdot \ldots \cdot W(c_{\theta(n)})$ for some permutation $\theta$, so we
require that these regular expressions be disjoint.

The third hypothesis says that the regular expression $K(c_1) \cdot K(c)
\cdot \ldots \cdot K(c) \cdot K(c_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \textit{permutation} QRE operates on
the language $K(c_1) \cdot K(c) \cdot \ldots \cdot K(c) \cdot K(c_n)$ which is
the kernel of the \textit{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

@@

\bcp{A question about variable names: If we use $R$ and $S$ for plain
  regular expressions, it seems inconsistent to use $c$ for QREs.  (And even
  more inconsistent to use $c$ and $c'$ or $c$ and $c_1$!) I suggest $P$ and
  $Q$ for QREs (with $Q$ as the default when we are only talking about
  one).}\sam{I made the notation consistent by just using $c_1$ and $c_2$. I
  personally feel that it is better not to use P and Q for QREs since we use p
  and q for QRE lenses, but if you disagree I will make the change.}\bcp{I
  agree that this is a problem.  But maybe the right fix is to choose
  different metavariables for QRE lenses.  Or to use capital letters other
  than $P$ and $Q$ for QREs.  I don't like $c_1$ and $c_2$, both for the
  lower-case letter and for the subscripts---it's not at all consistent with
our conventions elsewhere.}

Given regular expressions $R, S$ and equivalence relations
$\sim_R$ and $\sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, we\bcp{can we clarify that this is exactly Foster et al's
definition?}\sam{this is not Foster et al's definition. It's just a
loosening of the definition for bijective lens used in the Optician
paper}\bcp{OK, then we should say this}define
a {\em bijective quotient lens} $q :
R /{\sim_R} \Leftrightarrow S/{\sim_S}$ to be a pair of functions $q.\get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.\lput : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$, then $q.\lput(q_1.\get(r)) \sim_R r$ and
$q.\get(q_1.\lput(s)) \sim_R s$. Moreover, if $r \sim_R r'$, then $q.\get(r) =
q.\get(r')$, and if $s \sim_S s'$, then $q.\lput(s) = q.\lput(s')$.

@@

These laws are similar to the laws for bijective lenses, except that the
equality restriction in the bijective lens laws are loosened to allow for
equivalence relations. Also, the condition that $r \sim_R r'$ implies
$q.\get(r) = q.\get(r')$ ensures that the {\em get} function induced on
the equivalence classes if $\equiv_c$ is well-defined, and similarly for
the {\em put} function.

Now let $c_1$ and $c_2$ be QREs. Given a bijection $\ell : K(c_1)
\Leftrightarrow K(c_2)$, we may define a quotient lens $q : W(c_1)/\eqrel{c_1}
\Leftrightarrow W(c_2)/\eqrel{c_2}$ by $q.\get = \ell \circ \canonizer(c_1)
$ and $q.\lput = \ell_1^{-1} \circ \canonizer(c_2)$. In other words, the get
function of the $q$ first canonizes the source data using $\canonizer(c_1)$
and then applies $f$ to the result, while the put function of the $q$ first
canonizes the target data using $\canonizer(c_2)$ and the applies $f^{-1}$ to
the result. This is our approach in defining QRE lenses.

One potential pitfall with QRE lenses is that the combinators that
we define on them may not preserve this normal form. For example, the
composition $q_1 \; ; \; q_2$ of QRE lenses $q_1$ and $q_2$ may not be
equivalent to any quotient lens that has the canonical normal form of a
bijective lens with canonizers a the edges. Consequently, in order to guarantee
that every QRE lens $q_1$ is equivalent to a QRE lens $q_2$ which has this
canonical form, we need to identify a set $S$ of bijections that will ensure
that all QRE lenses is functionally equivalent to a QRE lens in canonical form.
Our choice is to use the class of \textit{bijective lenses}, which we studied in
\cite{optician} and has this desired property. The following part of the paper
briefly discusses bijective lenses.

\subsection{Bijective Lenses}
We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created using Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell_1 := \mathit{id} \; (R) \sep \const(s_1, s_2) \sep  \swap(\ell_1,
\ell_2) \sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \;
| \; \ell_1 \; ; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

>>>>

In our previous work in~\cite{optician}, we presented a solution to these
shortcomings of bijective lenses by developing a tool called {\em Optician}
which enables a programmer to synthesize a bijective lens from a pair of regular
expressions $(S, T)$ and a set of input-output example \cd{exs} pairs that
determines how the synthesized lens should behave on those examples. With {\em
Optician}, the programmer need only spend the effort of specifying the data
formats and the behaviour\bcp{Let's use American spellings: behavior} of the
lens on a few inputs in order to define a 
bijective lens. As a result defining bijective lenses becomes faster and less
error prone, since the programmer need only invoke the synthesis procedure
\lstinline{synth S <=> T using exs} in order to define a bijective lens rather
than manually derive the lens herself.

@@

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $c$ enables us to express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function\bcp{Why is it called ``canonizer'' rather
    than ``canonize''??} $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(c)$ and $K(c)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonizer(R) &=& id_{\mathcal{L}(R)} \\
\canonizer(\project{R}{s})(w) &=& s\\
\canonizer(\squash{c_1}{c_2}{f})(w) &=&
\begin{cases}
\canonizer(c_1)(f(w)) & \text{if } w \in \mathcal{L}(W(c_1))\\
\canonizer(c_2)(w) & \text{otherwise}
\end{cases}\\
\canonizer(\normalize{R}{R'}{f}) &=& f\\
\canonizer(c_1 \; ; \; c_2) &=& \canonizer(c_2) \circ \canonizer(c_1)\\
\canonizer(c_1 \cdot c_2) &=& \canonizer(c_1) \cdot \canonizer(c_2)\\
\canonizer(c_1 \sep c_2)(w) &=&
\begin{cases}
\canonizer(c_1)(w) & \text{if } w \in \mathcal{L}(W(c_1))\\
\canonizer(c_2)(w) & \text{if } w \in \mathcal{L}(W(c_2))\\
\end{cases}\\
\canonizer(c^*) &=& \canonizer(c)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \textit{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$c_i, c$ are placed in the canonical permutation which is the identity
permutation before $\canonizer(c_i)$ and $\canonizer(c)$ are applied
respectively.

Finally, Figure ~\ref{fig:relations} gives the inductive definition of the
equivalence relation $\eqrel{c}$. This is the set-theoretic semantics of QREs
as an equivalence relation on a regular language.

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_R \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{c_1}{c_2}{f}} \; w' &\iff& f(w) \equiv_{c_2} w'
\text{ or } w \equiv_{c_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{c_1 \; ; \; c_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(c_2)) \text{ such that } w \; \equiv_{c_1} \; k, \; w' \;
\equiv_{c_1} \; k', \text{ and } k \; \equiv_{c_2} \; k'\\
w \; \equiv_{c_1 \cdot c_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{c_1}
\; {r'}_1, \; r_2 \; \equiv{c_n} \; {r'}_2\\
w \; \equiv_{c_1 \sep c_2} \; w' &\iff& w \; \equiv_{c_1} \; w'
\text{ or } \; w \; \equiv_{c_2} \; w'\\
w \; \equiv_{c^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{c} \; {r'}_i
\\
w \; \equiv_{\perm{c_1, \ldots, c_n}{c}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{c} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Well Formed QREs}
\subsubsection{Completeness of {\em normalize}}
Before we present the full inference rules, for QREs, we discuss the {\em
normalize} QRE which is special in two main ways, the first of which has to do
with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

@@

Formally, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$. We say
that a regular expression $R$ is \textit{strongly unambiguous}\cite{Sippu1988}
if and only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1,
S_2$ strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \sep S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

@@

Figure~\ref{fig:qrerules} gives the inference rules for deriving well-formed
QREs.
\begin{figure}[tp!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\RightLabel{(Id)}
\UnaryInfC{$\wf{\mathit{id}(R)}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\RightLabel{(Project)}
\UnaryInfC{$\wf{\project{R}{s}}$}
\DisplayProof
\hskip 1.5em
\end{center}

@@

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$W(c_1) \cap W(c_2) = \varnothing$}
\RightLabel{(Union)}
\BinaryInfC{$\wf{c_1 \sep c_2}$}
\DisplayProof
\
\end{center}
\end{center}
\caption{Well-formed QREs}
\label{fig:qrerules}
\end{figure}
As we already mentioned, the unambiguity conditions come into play when defining
QREs using the regular combinators. For example, the (Concat) inference rule
for concatenation says that the concatenation $c_1 \cdot c_2$ of QREs $c_1$ and
$c_2$ is well formed only if the concatentions of $W(c_1)$ and $W(c_2)$, and
$K(c_1)$ and $K(c_2)$ are unambiguous.

The most complicated inference rule is the (Perm) rule for the
\textit{permutation} combinator. The second hypothesis for the
\textit{permutation} rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c_{\sigma(n)})$ and $W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot
W(c_{\theta(n)})$ must be disjoint. This is because any string that matches the
\textit{permutation} combinator matches the language $W(c_{\theta(1)}) \cdot
W(c) \cdot \ldots \cdot W(c_{\theta(n)})$ for some permutation $\theta$, so we
require that these regular expressions be disjoint.
%
The third hypothesis says that the regular expression $K(c_1) \cdot K(c)
\cdot \ldots \cdot K(c) \cdot K(c_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \textit{permutation} QRE operates on
the language $K(c_1) \cdot K(c) \cdot \ldots \cdot K(c) \cdot K(c_n)$ which is
the kernel of the \textit{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

@@

\bcp{A question about variable names: If we use $R$ and $S$ for plain
  regular expressions, it seems inconsistent to use $c$ for QREs.  (And even
  more inconsistent to use $c$ and $c'$ or $c$ and $c_1$!) I suggest $P$ and
  $Q$ for QREs (with $Q$ as the default when we are only talking about
  one).}\sam{I made the notation consistent by just using $c_1$ and $c_2$. I
  personally feel that it is better not to use P and Q for QREs since we use
  p and q for QRE lenses, but if you disagree I will make the change.}\bcp{I
  agree that this is suboptimal, but I don't think it's worse than what we
  have now.  Or maybe the right fix is to choose different metavariables for
  QRE lenses.  Or to use capital letters other than $P$ and $Q$ for QREs.
  In any case, I don't like $c_1$ and $c_2$, both for the lower-case letter
  and for the subscripts---it's not at all consistent with our conventions
  elsewhere.}

Given regular expressions $R, S$ and equivalence relations
$\sim_R$ and $\sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$,
we\bcp{can we clarify that this is exactly Foster et al's 
definition?}\sam{this is not Foster et al's definition. It's just a
loosening of the definition for bijective lens used in the Optician
paper}\bcp{OK, then we should say this}define
a {\em bijective quotient lens} $q :
R /{\sim_R} \Leftrightarrow S/{\sim_S}$ to be a pair of functions $q.\get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.\lput : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$, we have $q.\lput(q_1.\get(r)) \sim_R r$ and
$q.\get(q_1.\lput(s)) \sim_R s$. Moreover, if $r \sim_R r'$, then $q.\get(r) =
q.\get(r')$, and if $s \sim_S s'$, then $q.\lput(s) = q.\lput(s')$.

@@

These laws are similar to the laws for bijective lenses\bcp{we should state
  them explicitly somewhere in this paper and refer to them here by number},
except that the 
equality restrictions in the bijective lens laws are loosened to allow for
equivalence relations. Also, the condition that $r \sim_R r'$ implies
$q.\get(r) = q.\get(r')$ ensures that the {\em get} function induced on
the equivalence classes if $\equiv_c$ is well-defined, and similarly for
the {\em put} function.

Now let $c_1$ and $c_2$ be QREs. Given a bijection $\ell : K(c_1)
\Leftrightarrow K(c_2)$, we may define a quotient lens $q : W(c_1)/\eqrel{c_1}
\Leftrightarrow W(c_2)/\eqrel{c_2}$ by $q.\get = \ell \circ \canonizer(c_1)
$ and $q.\lput = \ell_1^{-1} \circ \canonizer(c_2)$. In other words, the get
function of $q$ first canonizes the source data using $\canonizer(c_1)$
and then applies $f$ to the result, while the put function of $q$ first
canonizes the target data using $\canonizer(c_2)$ and the applies $f^{-1}$ to
the result. This is our approach in defining QRE lenses.

One potential pitfall\bcp{Why is this a pitfall?  I.e., why do we care about
this norman form?  (The answer is synthesis, of course, but we haven't said
so here.)} with QRE lenses is that the combinators that
we define on them may not preserve this normal form\bcp{In what sense is
  this a ``normal form''?  What you've given above is just a definition:
  quotient lenses can be defined in such and such way.  You haven't said we
  want to consider it a normal form, or specified what process it is the
  normal form {\em of}.}. For example, the
composition $q_1 \; ; \; q_2$ of QRE lenses $q_1$ and $q_2$ may not be
equivalent to any quotient lens that has the canonical normal form of a
bijective lens with canonizers a the edges. Consequently, in order to guarantee
that every QRE lens $q_1$ is equivalent to a QRE lens $q_2$ in this
canonical form\bcp{This paragraph uses variables $q_1$ and $q_2$ twice, with
two different meanings.}, we need to identify a set $S$ of bijections that
will ensure 
that every QRE lens is functionally equivalent to a QRE lens in canonical
form.\bcp{This is not very clear, in part because it is not very precise:
  It is assuming, implicitly, that the class of QRE lenses is 
  parameterized on a particular set of bijections on kernels, and then
  asking what sets of bijections can be used.  This question makes sense,
  but the reader may not realize that this is what's being asked. } 
Our choice is to use the class of \textit{bijective lenses}, which we studied in
\cite{optician} and has this desired property. The following part of the paper
briefly discusses bijective lenses.

\subsection{Bijective Lenses}
\bcp{There's a persistent opportunity for confusion between the semantic
  space of {\em all} bijective lenses (perhaps between arbitrary structures,
  not just strings) and the rather specific bijections that are picked out
  by our combinators.  One way to help, I think, would be to use the phrase
  ``bijective string lenses'' instead of just ``bijective lenses.''}
We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created using Boomerang lens combinators.\bcp{Why
  do we choose Boomerang's set of combinators?  Does it have some formal
  property that we find attractive?  Or did it just happen to be lying around?}
The syntax for the language of bijective lenses is given by
$$\ell_1 := \mathit{id} \; (R) \sep \const(s_1, s_2) \sep  \swap(\ell_1,
\ell_2) \sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \;
| \; \ell_1 \; ; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.
